A Survey of Possible Exoskeleton Control Architectures and Algorithms\\ (Draft 2.2)

### 2015-09-22 control.tex

Note that we cannot change the apparent inertia of the exoskeleton without
some form of acceleration or force feedback.
With acceleration feedback we can make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding acceleration feedback:
\begin{equation}
\vtau = \vtau_{\invdynw} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
\label{eq:change-inertia1}
\end{equation}
sets the apparent inertia to $\mM_d$ if the dynamic models are perfect, and
the state estimation produces not only perfect positions and velocities but
also perfect acceleration estimates. In the case 
where the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}
In robotics this is usually considered a bad idea if $\mM_d$ is smaller (a complex
issue for a matrix) than $\mM$ at any $\vq$. This type of control is vulnerable
to unmodeled dynamics. However, in an exoskeleton, since we have not tried to
cancel the human's dynamics, the human stabilizes a possible unstable exoskeleton
(if the straps are tight enough).

@@

We implemented two kinds of Kalman filters for state estimation.
Low level state estimation is done by an extended Kalman filter that takes into
account joint position, actuator load measurements, actuator commands,
and the dynamic model of the robot,
and measured contact forces (feet and wrists)~\cite{}.
For higher level (overall) state estimation,
we developed a center of mass location and velocity estimator, and its
application in full body inverse dynamics control, fall detection
and fall prevention for humanoid robots and exoskeletons~\cite{}.

We use Linear
Inverted Pendulum Model (LIPM) dynamics with an offset to predict
the center of mass motion. This offset can be interpreted
as a modelling error on the center of mass position, or an
external force exerted on the center of mass of the robot,
or a combination of both. The center of mass estimator was
implemented on our Atlas humanoid robot. During the DARPA
Robotics Challenge Finals, it successfully saved our robot from
falling on two occasions, and made us the only competitive team
without a fall or need for a human rescue among all teams.

Unscented,
Particle filters

DISCUSS HOW TO HANDLE SENSOR NOISE

Estimating acceleration using double differentiation with low pass filtering
currently works well on real exoskeletons.
However, impacts, shock waves, and the fact that neither
the operator's body parts or the exoskeletons parts are rigid bodies and the
joints are not well defined suggest this approach can be improved with actual
acceleration measurements.
It is now possible to distribute MEMs IMUs throughout the exoskeleton to
improve both velocity and 
acceleration estimation~\cite{Xinjilefu-thesis}, which should improve
Sensitivity Amplification Control.

@@

\subsection{Full Atlas robot.}

@@

We need to be robust to parameter modeling errors such as the wrong mass
or moment of inertia for an exoskeleton part, variations in the operator
parameters, and variations in attachments to the exoskeleton and payload.
Examples of typically unmodeled dynamics (non-parametric modeling error) include:
backlash, joint play, structural deformation, flexible wires and hoses,
actuator dynamics, processing delay,
elastic operator tissue and straps, and 
loose or slack operator tissue and strap.

>>>>

Note that we cannot change the apparent inertia of the exoskeleton without
some form of acceleration or force feedback.
With acceleration feedback we can make the exoskeleton have a different inertia.
Setting $\ddot{\vq}_d$ to zero and adding acceleration feedback:
\begin{equation}
\vtau = \vtau_{\invdynw} - (\mM_d(\hat{\vq}) - \hat{\mM}(\hat{\vq})) \hat{\ddot{\vq}}
\label{eq:change-inertia1}
\end{equation}
sets the apparent inertia to $\mM_d$ if the dynamic models are perfect, and
state estimation produces not only perfect positions and velocities but
also perfect acceleration estimates. In the case 
where the operator directly applies torques at the
exoskeleton joints we get:
\begin{equation}
\mM_d (\vq) \ddot{\vq} = \vtau_o
\end{equation}
In robotics this is usually considered a bad idea if $\mM_d$ is smaller (a complex
issue for a matrix) than $\mM$ at any $\vq$. This type of control is vulnerable
to unmodeled dynamics. However, in an exoskeleton, since we have not tried to
cancel the human's dynamics, the human stabilizes a possible unstable exoskeleton
(if the straps are tight enough).

@@

We use extended Kalman Filtering (EKF) to perform state estimation.
Currently, using the full body dynamics (equation~\ref{eq:dynamics2})
and any addtional constraint equations for ground contacts (feet on the ground):
\begin{equation}
\mJ_w \ddot{\vq} + \dot{\mJ}_w \dot{\vq} = 0
\end{equation}
is too expensive to produce both a forward prediction and a linearized model
about the current state in the desired 1ms time step.
Therefore we approximate the dynamics by separating the center of mass dynamics
from the joint dynamics and use two separate extended Kalman filters:
a filter for the center of mass dynamics, and a filter for the joint dynamics.
We use an orthogonal (QR) decomposition to project motion into the orthogonal
complement of the contact Jacobian where the ``external'' dynamics (the dynamics
of the center of mass) are separate from the ``internal'' dynamics (the motion
of the joints).

Joint velocities are estimated using actuator position sensors, actuator load
sensors, actuator commands, a single high
quality IMU at the pelvis, multiple low quality (but cheap) MEMs IMUs
spread throughout the robot. We also take into account feet and hand
contact forces, which
are measured. We use the full internal dynamics as the process model,
assuming the root dynamics are known. See Chapter 3 of~\cite{Xinjelefu-thesis}.
The internal dynamics are linearized and covariances updated
by a separate process at a 100Hz rate.
The joint positions are predicted by integrating the estimated velocities.

For higher level (overall) state estimation,
we developed a center of mass location and velocity estimator, and its
application in full body inverse dynamics control, fall detection
and fall prevention for humanoid robots and exoskeletons~\cite{}.
The center of mass filter is modeled as a multiple model EKF with contact switching
to handle changes in support: single support, double support, and off the ground.
We use the robot joint sensing (kinematics) to create a virtual sensor that measures
center of mass position and velocity relative to a stance foot.
We use Linear
Inverted Pendulum Model (LIPM) dynamics with an offset to predict
the center of mass motion. This offset can be interpreted
as a modelling error on the center of mass position, or an
external force exerted on the center of mass of the robot,
or a combination of both. The center of mass estimator was
implemented on our Atlas humanoid robot. 
This estimator is especially helpful when compensating
for unplanned slow changing external forces applied at
unknown locations on the robot, which is quite likely when
operating in tight spaces. It also handles relatively small
dynamic forces well when walking, e.g. dragging a tether
or pushing through a spring loaded door. Thanks to
the estimator, very little tuning is done for our mass model.
During the DARPA
Robotics Challenge Finals
where no safety belay was
allowed, our early warning system
successfully saved our robot from
falling on two occasions, and made us the only competitive team
without a fall or need for a human rescue among all teams.

{\bf Safety Code: Fall Prediction:}
The most significant contribution of the external force estimator is that it
can detect when a large external force is being applied that might
push the robot over.
We compute a ``corrected capture point'' (CCP)~\cite{capture},
which is an offset to the current capture point.
The offset 
takes into account the estimated external force, represented as an offset
to the center of mass.
The corrected capture point getting close to the boundary of the polygon of
support warns the controller that the robot might fall if the external
force is maintained.
We can also compute the corrected capture point assuming that the
external force follows a known time course plus an estimated constant offset,
or steadily increases or decreases for a fixed time interval
based on an estimated derivative.
We assume the external force is due to the robot's actions, and not
due to external disturbances such as wind, a moving support platform
(although earthquakes do happen at the DRC location in California),
or external agents pushing on the robot.
The current
behavior is stopped and the robot ``frozen'' in place.

A derivation of the corrected capture point starts with LIPM dynamics
augmented with a true center of mass offset and a true external force:
\begin{equation}
\ddot{c} = (c + c_{offset} + f_{ext}*z/mg - cop)*g/z
= (c + \Delta - cop)*g/z
\end{equation}
where $c$ is the location of the center of mass projected on the ground
plane,
$cop$ is the center of pressure
location in that ground plane, and $\Delta$ is the sum of the true center of mass
offset from the modeled center of mass and any external horizontal force.
Our extended Kalman filter estimates $\hat{c}$, $\hat{\dot{c}}$, and
$\hat{\Delta}$,
taking into account the
current center of mass height $z$. We assume a constant center of mass height
in estimating the corrected capture point based on the estimated capture point
as described in~\cite{capture}:
\begin{equation}
\widehat{CCP} = \widehat{CP} + \hat{\Delta} 
= \hat{c} + \hat{\dot{c}}\hat{z}/g + \hat{\Delta}
\end{equation}

For dynamic walking, the corrected capture point goes beyond the
current single foot support polygon during swing phase and is captured by
the touchdown foot. One way to predict a fall is to use the
swing foot time to touchdown and predict if the current
capture point is within the support polygon of possible or desired
footholds. This approach is complicated because it depends
on the controller. Our simpler solution is to use a heuristic
that detects a fall only if the corrected capture point
is outside of the current support
polygon for a continuous period of time. The time threshold was set to
0.6 seconds after extensive testing. As soon as a fall is
predicted, there are several things the humanoid can do, such
as using angular momentum to maintain balance, or changing the
current or future foot placements.
We have implemented a simple
step recovery controller that works in single support where
the robot corrects the next foot placement and timing
using a simple heuristic to avoid self collision.
We would only step left if the robot is falling left, and similarly
for right steps. We avoid crossing the legs in using foot placement
to balance laterally.

\begin{figure}[t]
  \centering
  \hbox{}
      \includegraphics[height=5cm]{figs/door1-5.png}
%      \includegraphics[height=5cm]{figs/door2-1c.png}
      \hbox{}
      \includegraphics[height=5cm]{figs/door1-3.png}
      \hbox{}
      \includegraphics[height=5cm]{figs/door-stuckc2.png}
      \hbox{}
      \caption{Successful sidestepping through the door (left, middle)
        and the failure in the DRC rehearsal (right) in which the protective
        cage (black padding) for the head is against the white door frame.}
      \label{figDoor}
\end{figure}

\begin{figure}[t]
      \centering
      % \includegraphics[width=0.48\textwidth]{figs/door-traces.png}
\subfigure
{\includegraphics[width=0.48\textwidth]{figs/cop_x1.pdf} \label{subfig:ext_cop_x}}
%\subfigure
%{\includegraphics[width=0.48\textwidth]{figs/cop_y.pdf} \label{subfig:ext_cop_y}}
\subfigure
{\includegraphics[width=0.48\textwidth]{figs/ext_f1.pdf} \label{subfig:ext_f}}
\caption{
Atlas was caught on the door frame when sidestepping through it during
the DRC rehearsal.
The walking controller delayed liftoff and remained in double
support when the external force estimator detected a large change
in the estimated external force in the robot's sideways
direction ($F_x$, through the door).
The single support phase is shown by the shaded area, and the black dashed
lines indicates the planned liftoff time.
The estimated CoM is the sum of the model CoM and the estimated CoM offset.
}
\label{figDoorError}
\end{figure}

{\bf Results of Safety Code: Robot caught on door frame:}
In the DRC rehearsal, the robot was caught on
the door frame when sidestepping though (Figure~\ref{figDoor}).
The walking
controller detected an anomaly in the estimated external force in the
sideways direction ($F_x$),
delayed liftoff and remained in double support, and stopped the
current behavior
to allow for manual recovery (Figure~\ref{figDoorError}).

{\bf Results of Safety Code: Manipulation Error:}
For the manipulation controller, the robot is always assumed
to be in double support, and the support polygon
is computed by finding the convex hull of the foot corner
points (light green in Figure~\ref{figDrillCOP}),
computed using forward kinematics.
To prevent the robot from falling during
manipulation, we require the corrected capture point to be within a
subset of the support polygon called the safe region, 
(dark green in Figure~\ref{figDrillCOP}),
When the corrected capture point escapes the safe region,
a freeze signal is sent to the manipulation controller, and it
clears all currently executing joint trajectories and freezes the robot
at the current pose, with the balance controller still running.

During our second
run in the Finals, our right electric forearm mechanically failed
when the cutting motion was initiated for the drill task. The
uncontrolled forearm wedged the drill into the wall and
pushed the robot backwards.
The controller stopped the behavior (a freeze),
and saved the robot from falling (Figure~\ref{figDrillCOP}).
The operator was then able to
recover from an otherwise catastrophic scenario.

The time plot in Figure~\ref{figDrillCOP} shows candidate fall predictors
during this event. We can eliminate some candidate fall predictors
easily. The center of mass (CoM) (and a ``corrected'' CoM (not shown))
usually provide a fall warning too late, because the CoM velocity is not
included. The capture point (CP) does not include information about
external forces. The center of pressure (CoP) is 
noisy and gives too many false alarms. It can warn of 
foot tipping, but it is less reliable about warning about robot falling,
which is not the same thing as foot tipping in a force controlled robot
or if there are non-foot contacts and external
forces. In this plot, we see that the CoP moves away from the safe region
during recovery, predicting that the robot is falling again, while the
corrected capture point (CCP) moves towards the interior of the safe region.

\begin{figure}[t]
      \centering
      \includegraphics[height=30mm]{figs/near-fall2c.png}
      \includegraphics[height=30mm]{figs/drill-cop1.png}
      \includegraphics[height=30mm]{figs/drill-cop2.png}
      \includegraphics[height=30mm]{figs/drill-cop3.png}\\
\includegraphics[width=0.48\textwidth]{figs/cp_time_xes.pdf}
      \caption{Top Left: Robot posture after 
error detection.
Top Right: Black trace: The corrected capture point (CCP)
up to the error detection,
Red trace: The CCP during the ``frozen'' period, and Blue trace: The CCP moves
back to the center of the polygon of support during manual recovery.
Bottom: Plots of candidate fall predictors in the fore/aft direction
during this event. The black vertical dashed lines mark the freeze time and
the start of manual recovery.}
      \label{figDrillCOP}
\end{figure}

Unscented and particle filters could be used instead of extended
Kalman filtering. It may well be the case that unscented filtering
outperforms extended Kalman filtering.

\subsection{Current state estimation approaches in exoskeletons}

We see that sensitivity amplification control uses acceleration as an input.
Estimating acceleration using double differentiation with low pass filtering
currently works well on real exoskeletons.
However, impacts, shock waves, and the fact that neither
the operator's body parts or the exoskeletons parts are rigid bodies and the
joints are not well defined suggest this approach can be improved with actual
acceleration measurements.
It is now possible to distribute MEMs IMUs throughout the exoskeleton to
improve both velocity and 
acceleration estimation~\cite{Xinjilefu-thesis}, which should improve
Sensitivity Amplification Control.

@@

@@

We need to be robust to parameter modeling errors such as the wrong mass,
center of mass location, joint location,
or moment of inertia for an exoskeleton part, variations in the operator
parameters, and variations in attachments to the exoskeleton and payload.
Examples of typically unmodeled dynamics (non-parametric modeling error) include:
backlash, joint play, structural deformation, flexible wires and hoses,
actuator dynamics, processing delay,
elastic operator tissue and straps, and 
loose or slack operator tissue and strap.
